---
title: "GIS Tutorial"
author: "Rachel Saidi"
date: "10/20/2020"
---


# This tutorial is based on Computerworld's How to Make a Map with R 

In 10 (fairly) Easy Steps https://www.computerworld.com/article/3038270/data-analytics/create-maps-in-r-in-10-fairly-easy-steps.html by Sharon Machlis sharon_machlis@idg.com

```{r}
# Set various values needed, including names of files and FIPS codes for New Hampshire and South Carolina
nhdatafilecsv <- "NHD2016.csv"
usshapefile <- "cb_2014_us_county_5m/cb_2014_us_county_5m.shp"
nhfipscode <- "33"
scdatafile <- "SCGOP2016.csv"
scfipscode <- "45"
```

### Run any of the install.packages() commands below for packages that are not yet on your system
 install.packages("tmap")
 install.packages("tmaptools")
 install.packages("leaflet")
 install.packages("scales")
 install.packages("leaflet.extras")
 install.packages("rio")
 install.packages("htmlwidgets")
 install.packages("sf")
 install.packages("dplyr")
 install.packages("sp")

## Load the tmap, tmaptools, and leaflet packages into your working session:
 
```{r}
library(tidyverse)
library(tmap)
library(tmaptools)
library(leaflet)
library(sf)
library(leaflet.extras)
library(rio)
library(sp)
```

 
## Step 1: Read in the NH election results file:

 
```{r}
setwd("C:/Users/rsaidi/Dropbox/Rachel/MontColl/Datasets/GIS")
nhdata <- import(nhdatafilecsv)
```

If you have any problems with this, there is also a CSV version of the file -- sometimes reading Excel between Mac and Windows can be tricky. Try

nhdata <- rio::import(nhdatafilecsv)

## Eliminate columns for minor candidates and just use County, Clinton and Sanders columns:
 
```{r}
nhdata <- nhdata[,c("County", "Clinton", "Sanders")]
```

## Step 2: Decide what data to map

### Add columns for percents and margins:

```{r}
nhdata$SandersMarginVotes <- nhdata$Sanders - nhdata$Clinton
nhdata$SandersPct <- (nhdata$Sanders) / (nhdata$Sanders + nhdata$Clinton) 
# Will use formatting later to multiply by a hundred
nhdata$ClintonPct <- (nhdata$Clinton) / (nhdata$Sanders + nhdata$Clinton)
nhdata$SandersMarginPctgPoints <- nhdata$SandersPct - nhdata$ClintonPct
```

## Step 3: Get geographic data files

### Read in the shapefile for US states and counties:

* If libraries with raster and rgdal don't work (see next chunk), try library(sf) with the command st_read

* All these options are here and should help you get the qtm command in the next chunk


```{r}
#install.packages("raster")
#install.packages("rgdal")
library(raster)
library(rgdal)
setwd("C:/Users/rsaidi/Dropbox/Rachel/MontColl/Datasets/GIS")
usgeo <- shapefile("cb_2014_us_county_5m/cb_2014_us_county_5m.shp")
```

## Do a quick plot (qtm stands for quick thematic map) of the shapefile and check its structure:

```{r }
tmap_options(check.and.fix = TRUE)
qtm(usgeo)
# (pause to wait for map to render, may take a few seconds)
```

## Subset just the NH data from the US shapefile

```{r}
nhgeo <- usgeo[usgeo$STATEFP==nhfipscode,]
```

## tmap test plot of the New Hampshire data

```{r}
qtm(nhgeo)
```

## Match the two dataset variable headers for county (One is NAME and the other is County)

```{r}
# Check if county names are in the same format in both files
str(nhgeo$NAME)
str(nhdata$County)
nhgeo$NAME  # they are the same
```

## Order each data set by county name (alphabetically)

```{r message = FALSE}
nhgeo <- nhgeo[order(nhgeo$NAME),]
nhdata <- nhdata[order(nhdata$County),]
if (identical(nhgeo$NAME,nhdata$County)) {
nhmap <- merge(nhgeo, nhdata, by.x = "NAME", by.y = "County")  # Merge geo and vote datasets
} else {stop}
```

## Step 4: Merge geo data with results data using the merge function

```{r}
# See the new data structure with
head(nhmap)
```


## Step 5: Create a static map with tmap's qtm() function:

```{r}
qtm(nhmap, "SandersMarginVotes")
qtm(nhmap, "SandersMarginPctgPoints")
```

## For more control over look and feel, use the tm_shape() function:

```{r}
tm_shape(nhmap) +
  tm_fill("SandersMarginPctgPoints", title="Sanders Margin, Total Votes", palette = "PRGn") +
  tm_borders(alpha=.5) +
  tm_text("NAME", size=0.8)
```

## Same code as above, but store the static map in a variable, and change the theme to "classic" style:

```{r}
nhstaticmap <- tm_shape(nhmap) +
tm_fill("SandersMarginPctgPoints", title="Sanders Margin, Total Votes", palette = "PRGn") + #I like viridis
tm_borders(alpha=.5) +
tm_text("NAME", size=0.8) + 
tm_style("classic")
```

## View the map

```{r}
nhstaticmap
```


## Part 6 

Next up: Code for a basic interactive map, this time for Clinton percentages in NH

## Create a palette

```{r}
clintonPalette <- colorNumeric(palette = "Blues", domain=nhmap$ClintonPct)  # color palette for Clinton

```

## and a pop-up window


```{r}
library(scales)
nhpopup <- paste0("County: ", nhmap$NAME, "<br /><br /> Sanders:  ", percent(nhmap$SandersPct), " Clinton: ", percent(nhmap$ClintonPct))  # popup content
```

## Step 7: Now generate the interactive map:

For more information on CRS (coordinate reference systems) projection, see this document:  https://rspatial.org/raster/spatial/6-crs.html


```{r}
# add the appropriate projection, WGS84
nhmap_projected <- spTransform(nhmap, "+proj=longlat +datum=WGS84")  
leaflet(nhmap_projected) %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(stroke=FALSE, 
              smoothFactor = 0.2, 
              fillOpacity = .8, 
              popup=nhpopup,
              color= ~clintonPalette(nhmap$ClintonPct))

```

## South Carolina data

```{r}
setwd("C:/Users/rsaidi/Dropbox/Rachel/MontColl/Datasets/GIS")
scdata <- rio::import(scdatafile)
```

## South Carolina shapefile and Quick plot of scgeo SC geospatial object:

```{r}
scgeo <- usgeo[usgeo@data$STATEFP=="45",]
qtm(scgeo)
```


## Add a column with percent of votes for each candidate. Candidates are in columns 2-7:


```{r}
candidates <- colnames(scdata[2:7])
for(i in 2:7){
  j = i + 7
  temp <- scdata[[i]] / scdata$Total
  scdata[[j]] <- temp
  colnames(scdata)[j] <- paste0(colnames(scdata)[i], "Pct")
}  
winner <- colnames(scdata[2:7])
```

## Get winner in each precinct

```{r}
for(i in 1:nrow(scdata)){
  scdata$winner[i] <- names(which.max(scdata[i,2:7]))
}
```

## Import spreadsheet with percent of adult population holding at least a 4-yr college degree


```{r}
setwd("C:/Users/rsaidi/Dropbox/Rachel/MontColl/Datasets/GIS")
sced <- rio::import("SCdegree.xlsx")
```

## Check if county names are in the same format in both files

```{r}
# Change the county names to plain characters in scgeo:
scgeo$NAME <- as.character(scgeo$NAME)
# Order each data set by county name
scgeo <- scgeo[order(scgeo$NAME),]
scdata <- scdata[order(scdata$County),]
# check name and county columns are identical
if (identical(scgeo$NAME,scdata$County )) {
scmap <- merge(scgeo, scdata, by.x = "NAME", by.y = "County")
} else {stop}

str(scgeo$NAME)
str(scdata$County)
```


## Instead of just coloring the winner, let's color by strength of win with multiple layers


```{r}
# Use same intensity for all - get minimum and maximum for the top 3 combined
minpct <- min(c(scdata$`Donald J TrumpPct`, scdata$`Marco RubioPct`, scdata$`Ted CruzPct`))
maxpct <- max(c(scdata$`Donald J TrumpPct`, scdata$`Marco RubioPct`, scdata$`Ted CruzPct`))
```

## Create leaflet palettes for each layer of the map:

```{r}
trumpPalette <- colorNumeric(palette = "Purples", domain=c(minpct, maxpct))
rubioPalette <- colorNumeric(palette = "Reds", domain = c(minpct, maxpct))
cruzPalette <- colorNumeric(palette = "Oranges", domain = c(minpct, maxpct))

winnerPalette <- colorFactor(palette=c("#984ea3", "#e41a1c"), domain = scmap$winner)
edPalette <- colorNumeric(palette = "Blues", domain=scmap$PctCollegeDegree)
```

## Create a pop-up:

```{r}
scpopup <- paste0("<b>County: ", scmap$NAME, "<br />Winner: ", scmap$winner, "</b><br /><br />Trump: ", percent(scmap$`Donald J TrumpPct`), "<br />Rubio: ", percent(scmap$`Marco RubioPct`), "<br />Cruz: ", percent(scmap$`Ted CruzPct`), "<br /><br />Pct w college ed: ", sced$PctCollegeDegree, "% vs state-wide avg of 25%")
```

## Add the projection we know from the NH map we'll need for this data on a Leaflet map:

```{r}
scmap <- sp::spTransform(scmap, "+proj=longlat +datum=WGS84")
```

## Basic interactive map showing winner in each county:

```{r}
leaflet(scmap) %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(stroke=TRUE,
              weight=1,
              smoothFactor = 0.2,
              fillOpacity = .75,
              popup=scpopup, 
              color= ~winnerPalette(scmap$winner),
              group="Winners" ) %>%
    addLegend(position="bottomleft", colors=c("#984ea3", "#e41a1c"), labels=c("Trump", "Rubio"))
```

## Put top 3 candidates in their own layers and add education layer, store in scGOPmap2 variable

```{r}
scGOPmap <- leaflet(scmap) %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(stroke=TRUE,
              weight=1,
              smoothFactor = 0.2,
              fillOpacity = .75,
              popup=scpopup, 
              color= ~winnerPalette(scmap$winner),
              group="Winners"  ) %>% 
    addLegend(position="bottomleft", colors=c("#984ea3", "#e41a1c"), labels=c("Trump", "Rubio"))  %>%

  addPolygons(stroke=TRUE,
     weight=1,
     smoothFactor = 0.2, 
     fillOpacity = .75, 
     popup=scpopup, 
     color= ~trumpPalette(scmap$`Donald J TrumpPct`),
     group="Trump") %>%

  addPolygons(stroke=TRUE,
              weight=1,
              smoothFactor = 0.2, 
              fillOpacity = .75, 
              popup=scpopup, 
              color= ~rubioPalette(scmap$`Marco RubioPct`),
              group="Rubio") %>%

  addPolygons(stroke=TRUE,
              weight=1,
              smoothFactor = 0.2, 
              fillOpacity = .75, 
              popup=scpopup, 
              color= ~cruzPalette(scmap$`Ted CruzPct`),
              group="Cruz") %>%

  addPolygons(stroke=TRUE,
              weight=1,
              smoothFactor = 0.2, 
              fillOpacity = .75, 
              popup=scpopup, 
              color= ~edPalette(sced$PctCollegeDegree), #this data is in the sced table, not scmaps
              group="College degs") %>%

  addLayersControl(
      baseGroups=c("Winners", "Trump", "Rubio", "Cruz", "College degs"),
      position = "bottomleft",
      options = layersControlOptions(collapsed = FALSE))

# Now display the map
scGOPmap
```


