---
title: "Healthy Cities GIS Assignment"
author: "R Saidi"
format: pdf
editor: visual
---

## Load the libraries and set the working directory

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(tidyr)
setwd("C:/Users/rsaidi/Dropbox/Rachel/MontColl/Datasets/Datasets")
cities500 <- read_csv("500CitiesLocalHealthIndicators.cdc.csv")
```

## The GeoLocation variable has (lat, long) format

Split GeoLocation (lat, long) into two columns: lat and long

```{r}
latlong <- cities500|>
  mutate(GeoLocation = str_replace_all(GeoLocation, "[()]", ""))|>
  separate(GeoLocation, into = c("lat", "long"), sep = ",", convert = TRUE)
head(latlong)
```

## Filter the dataset

Remove the StateDesc that includes the United Sates, select **Prevention** as the category (of interest), filter for only measuring **crude prevalence** and select only **2017**.

```{r}
latlong_clean <- latlong |>
  filter(StateDesc != "United States") |>
  filter(Category == "Prevention") |>
  filter(Data_Value_Type == "Crude prevalence") |>
  filter(Year == 2017)
head(latlong_clean)
```

## What variables are included? (can any of them be removed?)

```{r}
names(latlong_clean)
```

## Remove the variables that will not be used in the assignment

```{r}
prevention <- latlong_clean |>
  select(-DataSource,-Data_Value_Unit, -DataValueTypeID, -Low_Confidence_Limit, -High_Confidence_Limit, -Data_Value_Footnote_Symbol, -Data_Value_Footnote)
head(prevention)
md <- prevention |>
  filter(StateAbbr=="MD")
head(md)
```

The new dataset "Prevention" is a manageable dataset now.

# For your assignment, work with the cleaned "Prevention" dataset

### 1. Once you run the above code, filter this dataset one more time for any particular subset.

Filter chunk here

```{r}

```

### 2. Based on the GIS tutorial (Japan earthquakes), create one plot about something in your subsetted dataset.

First plot chunk here

```{r}

```

### 3. Now create a map of your subsetted dataset.

First map chunk here

```{r}

```

### 4. Refine your map to include a mousover tooltip

Refined map chunk here

```{r}

```

### 5. Write a paragraph

In a paragraph, describe the plots you created and what they show.
