---
title: "Hate Crimes in NY from 2010-2016"
author: "Rachel Saidi"
date: "May 5, 2020"
format: html
editor: 
  markdown: 
    wrap: 72
---


## Hate Crimes Dataset

This dataset looks at all types of hate crimes in New York counties by
the type of hate crime from 2010 to 2016.

My caveat:

### Flawed hate crime data collection - we should know how the data was collected

(Nathan Yau of Flowing Data, Dec 5, 2017)

Data can provide you with important information, but when the collection
process is flawed, there's not much you can do. Ken Schwencke, reporting
for ProPublica, researched the tiered system that the FBI relies on to
gather hate crime data for the United States:

"Under a federal law passed in 1990, the FBI is required to track and
tabulate crimes in which there was 'manifest evidence of prejudice'
against a host of protected groups, regardless of differences in how
state laws define who's protected. The FBI, in turn, relies on local law
enforcement agencies to collect and submit this data, but can't compel
them to do so."

This is a link to the ProPublica Article:
https://www.propublica.org/article/why-america-fails-at-gathering-hate-crime-statistics

Here is a data visualization of where hate crimes do NOT get reported
around the country (Ken Schwencke, 2017):
https://projects.propublica.org/graphics/hatecrime-map

### So now we know that there is possible bias in the dataset, what can we do with it?

```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
setwd("C:/Users/rsaidi/Dropbox/Rachel/MontColl/Datasets/Datasets")
hatecrimes <- read_csv("hateCrimes2010.csv")
```


## Clean up the data:

### Make all headers lowercase and remove spaces

After cleaning up the variable names, look at the structure of the data.
Since there are 44 variables in this dataset, you can use "summary" to decide which hate crimes to focus on. In the output of "summary", look at the min/max values. Some have a max-vale of 1.

```{r}
names(hatecrimes) <- tolower(names(hatecrimes))
names(hatecrimes) <- gsub(" ","",names(hatecrimes))
head(hatecrimes)
```

## Select only certain hate-crimes

```{r}
summary(hatecrimes)
```


I decided I would only look at the hate-crime types with a max number of 9 or more. That way I can focus on the most prominent types of hate-crimes.

```{r}
hatecrimes2 <- hatecrimes |>
  select(county, year, 'anti-black', 'anti-white', 'anti-jewish', 'anti-catholic','anti-age*','anti-islamic(muslim)', `anti-multi-religiousgroups`, 'anti-gaymale', 'anti-hispanic', `anti-otherethnicity/nationalorigin`) |>
  group_by(county, year)
head(hatecrimes2)
```

## Check the dimensions and the summary to make sure no missing values

Also check the dimensions to count how many variables remain

```{r}
dim(hatecrimes2)
# There are currently 12 variables with 423 rows.
summary(hatecrimes2)
```

## Convert from wide to long format

Look at each set of hate-crimes for each type for each year. Convert the dataset from wide to long with the pivot_longer function. It will take each column's hate-crime type combine them all into one column called "victim_cat". Then each cell count will go into the new column, "crimecount". 

Finally, we are only doing this for the quantitiative variables, which are in columns 3 - 10. Note the command facet_wrap requires (\~) before "victim_cat".

```{r}
hatelong <- hatecrimes2 |> 
    pivot_longer(
        cols = 3:12,
        names_to = "victim_cat",
        values_to = "crimecount")
```

## Now use the long format to create a facet plot

```{r}
hatecrimplot <-hatelong |> 
  ggplot(aes(year, crimecount))+
  geom_point()+
  aes(color = victim_cat)+
  facet_wrap(~victim_cat)
hatecrimplot
```

## Look deeper into crimes against blacks, gay males, and jews

From the facet_wrap plot above, anti-black, anti-gay males, and
anti-jewish categories seem to have highest rates of offenses reported.
Filter out just for those 3 crimes.

```{r}
hatenew <- hatelong |>
  filter( victim_cat %in% c("anti-black", "anti-jewish", "anti-gaymale"))|>
  group_by(year, county) |>
  arrange(desc(crimecount))
hatenew
```

## Plot these three types of hate crimes together

Use the following commands to finalize your barplot: - position =
"dodge" makes side-by-side bars, rather than stacked bars - stat =
"identity" allows you to plot each set of bars for each year between
2010 and 2016 - ggtitle gives the plot a title - labs gives a title to
the legend

```{r}
plot2 <- hatenew |>
  ggplot() +
  geom_bar(aes(x=year, y=crimecount, fill = victim_cat),
      position = "dodge", stat = "identity") +
  labs(fill = "Hate Crime Type",
       y = "Number of Hate Crime Incidents",
       title = "Hate Crime Type in NY Counties Between 2010-2016",
       caption = "Source: NY State Division of Criminal Justice Services")
plot2
```

We can see that hate crimes against jews spiked in 2012. All other years
were relatively consistent with a slight upward trend. There was also an
upward trend in hate crimes against gay males. Finally, there appears to
be a downward trend in hate crimes against blacks during this period.

## What about the counties?

I have not dealt with the counties, but I think that is the next place
to explore. I can make bar graphs by county instead of by year.

```{r}
plot3 <- hatenew |>
  ggplot() +
  geom_bar(aes(x=county, y=crimecount, fill = victim_cat),
      position = "dodge", stat = "identity") +
  labs(fill = "Hate Crime Type",
       y = "Number of Hate Crime Incidents",
       title = "Hate Crime Type in NY Counties Between 2010-2016",
       caption = "Source: NY State Division of Criminal Justice Services")
plot3
```

## So many counties

There are too many counties for this plot to make sense, but maybe we
can just look at the 5 counties with the highest number of incidents. -
use "group_by" to group each row by counties - use summarize to get the
total sum of incidents by county - use arrange(desc) to arrange those
sums of total incidents by counties in descending order. 

```{r}
counties <- hatenew |>
  group_by(year, county)|>
  summarize(sum = sum(crimecount)) |>
  arrange(desc(sum))
counties
```

## Top 5

To list the 5 counties with the highest total incidents, change group_by to: group_by(county), then use slice_max(order_by = sum, n=5) to
list the 5 counties with highest total incidents

```{r}
counties2 <- hatenew |>
  group_by(county)|>
  summarize(sum = sum(crimecount)) |>
  slice_max(order_by = sum, n=5)
counties2
```

Finally, create the barplot above, but only for the 5 counties in 2012
with the highest incidents of hate-crimes. The command "labs" is nice,
because you can get a title, subtitle, y-axis label, and legend title,
all in one command.

```{r}
plot4 <- hatenew |>
  filter(county %in% c("Kings", "New York", "Suffolk", "Nassau", "Queens")) |>
  ggplot() +
  geom_bar(aes(x=county, y=crimecount, fill = victim_cat),
      position = "dodge", stat = "identity") +
  labs(y = "Number of Hate Crime Incidents",
       title = "5 Counties in NY with Highest Incidents of Hate Crimes",
       subtitle = "Between 2010-2016", 
       fill = "Hate Crime Type",
      caption = "Source: NY State Division of Criminal Justice Services")
plot4
```

## How would calculations be affected by looking at hate crimes in counties per year by population densities?

Bring in census data for populations of New York counties. These are
estimates from the 2010 census.

```{r}
setwd("C:/Users/rsaidi/Dropbox/Rachel/MontColl/Datasets/Datasets")
nypop <- read_csv("newyorkpopulation.csv")
```

## Clean the county name to match the other dataset

Rename the variable "Geography" as "county" so that it matches in the
other dataset.

```{r}
nypop$Geography <- gsub(" , New York", "", nypop$Geography)
nypop$Geography <- gsub("County", "", nypop$Geography)
nypoplong <- nypop |>
  rename(county = Geography) |>
  gather("year", "population", 2:8) 
nypoplong$year <- as.double(nypoplong$year)
head(nypoplong)
```

## Focus on 2012

Since 2012 had the highest counts of hate crimes, let's look at the
populations of the counties in 2012.

Clean the nypoplong12 variable, county, so that matches the counties12
variable by Cutting off the ", New York" portion of the county listing

```{r}
nypoplong12 <- nypoplong |>
  filter(year == 2012) |>
  arrange(desc(population)) |>
  head(10)
nypoplong12$county<-gsub(" , New York","",nypoplong12$county)
nypoplong12
```

Not surprisingly, 4/5 of the counties with the highest populations also
were listed in the counties with the highest number of hate crimes. Only
the Bronx, which has the fifth highest population is not in the list
with the highest number of total hate crimes over the period from 2010
to 2016.

#### Recall the total hate crime counts:

Kings 713
New York 459
Suffolk 360
Nassau 298
Queens 235

## Filter hate crimes just for 2012 as well

```{r}
counties12 <- counties |>
  filter(year == 2012) |>
  arrange(desc(sum)) 
counties12
```

## Join the Hate Crimes data with NY population data for 2012

```{r}
datajoin <- counties12 |>
  full_join(nypoplong12, by=c("county", "year"))
datajoin
```

## Calculate the rate of incidents per 100,000. Then arrange in descending order

```{r}
datajoinrate <- datajoin |>
  mutate(rate = sum/population*100000) |>
  arrange(desc(rate))
datajoinrate
```

Notice that the highest rates of hate crimes in 2012 happened in:

```{r}
dt <- datajoinrate[,c("county","rate")]
dt
```

But the highest populated counties were: Kings (Brooklyn), Queens, New
York, Suffolk (Long Island), Bronx, and Nassau. They do not correspond
directly, though they are similar, to the counties with highest rates of
hate crimes.

## Follow Up

### Aggregating some of the categories

```{r}

aggregategroups <- hatecrimes |>
  pivot_longer(
    cols = 4:44,
    names_to = "victim_cat",
    values_to = "crimecount"
  )
unique(aggregategroups$victim_cat)
```

```{r}
aggregategroups <- aggregategroups |>
  mutate(group = case_when(
    victim_cat %in% c("anti-transgender", "anti-gayfemale", "anti-gendervictim_catendityexpression", "anti-gaymale", "anti-gay(maleandfemale", "anti-bisexual") ~ "anti-lgbtq",
    victim_cat %in% c("anti-multi-racialgroups", "anti-jewish", "anti-protestant", "anti-multi-religousgroups", "anti-religiouspracticegenerally", "anti-buddhist", "anti-hindu", "anti-mormon", "anti-sikh", "anti-catholic", "anti-islamic(muslim)", "anti-atheism/agnosticism", "anti-otherreligion", "anti-easternorthodox(greek,russian,etc.)", "anti-jehovahswitness", "anti-otherchristian") ~ "anti-religion", 
    victim_cat %in% c("anti-asian", "anti-arab", "anti-non-hispanic", "anti-white", "anti-americanindian/alaskannative", "anti-nativehawaiian/pacificislander", "anti-otherrace", "anti-hispanic", "anti-otherethnicity/nationalorigin") ~ "anti-ethnicity",
    victim_cat %in% c("anti-physicaldisability", "anti-mentaldisability") ~ "anti-disability",
    victim_cat %in% c("anti-female", "anti-male") ~ "anti-gender",
    TRUE ~ "others"))
aggregategroups
```

### or create subset with just lgbtq

```{r}
lgbtq <- hatecrimes |>
  pivot_longer(
      cols = 4:44,
      names_to = "victim_cat",
      values_to = "crimecount") |>
filter(victim_cat %in% c("anti-transgender", "anti-gayfemale", "anti-gendervictim_catendityexpression", "anti-gaymale", "anti-gay(maleandfemale", "anti-bisexual"))
lgbtq
```

## So what does all of this mean?

### Important Findings:

1.  I wonder what the data would look like if there was a universally
    accepted requirement for this type of data collection.

2.  The Bronx appears to have much lower than expected victim counts of hate
    crimes relative to its population density in comparison to other NY
    counties.

3.  In Kings County, NY (which is home to Brooklyn; according to
    Wikipedia, it is New York's most populous borough and the second
    most densly populated county in the US) in 2012, there was a spike
    in hate crimes against jews.

4.  All of these findings are corroborated in Hate Crime in New York
    State 2012 Annual Report:
    https://www.criminaljustice.ny.gov/crimnet/ojsa/hate-crime-in-nys-2012-annual-report.pdf

# Thank you!!! Questions?

------------------------------------------------------------------------

------------------------------------------------------------------------
