---
title: "Webscrape Wikipedia Formula One Drivers "
format: docx
---

## Load the libraries

```{r}
library(tidyverse)
library(rvest)
```

## Link to the URL you intend to scrape

HTTP GET request

https://en.wikipedia.org/wiki/List_of_Formula_One_drivers

```{r}
url <-"https://en.wikipedia.org/wiki/List_of_Formula_One_drivers"
webpage <- read_html(url)
class(webpage)
```

## Extract all tables

```{r}
tables <- webpage |>
  html_nodes("table.wikitable") |>
  html_table(fill = TRUE)

# How many tables did we find?
length(tables)
```

*There are three tables.*

## Inspect the size of the first table

```{r}
# Print the number of rows in each table
sapply(tables, nrow)
```

## Select the Drivers Table (the second table)

```{r}
drivers_raw <- tables[[2]]

# Preview
glimpse(drivers_raw)
head(drivers_raw)
```

## Clean the table

```{r}
names(drivers_raw) <- tolower(names(drivers_raw))
names(drivers_raw) <- gsub(" ", "_", names(drivers_raw))
names(drivers_raw)
```



## Use the Formula_one Drivers data frame to create your own visualization

Be sure it has a title, axes labels, caption for the data source (the website url). Make sure you include at least 3 variables in your plot that show: colors, sizes or other distinguishing elements.



*******************************************************************************

